---
title: "Tutorial 7: Bayesian Updating"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(shiny)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
tutorial_options(exercise.eval = FALSE)

my_reveal_time <- "2026-06-20 15:00:00"

theta_grid <- seq(0, 1, length.out = 101)

is_server_context <- function(.envir) {
  inherits(.envir$input, "reactivevalues") &
    inherits(.envir$output, "shinyoutput") &
    inherits(.envir$session, "ShinySession")
}

check_server_context <- function(.envir) {
  if (!is_server_context(.envir)) {
    calling_func <- deparse(sys.calls()[[sys.nframe() - 1]])
    err <- paste0("Function `", calling_func, "` must be called from an Rmd chunk where `context = \"server\"`")
    stop(err, call. = FALSE)
  }
}

encoder_logic <- function(strip_output = FALSE) {
  p <- parent.frame()
  check_server_context(p)
  assign("strip_output", strip_output, envir = p)
  local(
    {
      encoded_txt <- shiny::eventReactive(
        input$hash_generate,
        {
          state <- learnr:::get_tutorial_state()
          shiny::validate(shiny::need(length(state) > 0, "No progress yet."))
          shiny::validate(shiny::need(nchar(input$name) > 0, "No name entered."))
          shiny::validate(shiny::need(nchar(input$studentID) > 0, "Please enter your student ID"))
          user_state <- purrr::map_dfr(state, identity, .id = "label")
          user_state <- dplyr::group_by(user_state, label, type, correct)
          user_state <- dplyr::summarize(
            user_state,
            answer = list(answer),
            timestamp = dplyr::first(timestamp),
            .groups = "drop"
          )
          user_state <- dplyr::relocate(user_state, correct, .before = timestamp)
          user_info <- tibble(
            label = c("student_name", "student_id"),
            type = "identifier",
            answer = as.list(c(input$name, input$studentID)),
            timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "UTC")
          )
          learnrhash::encode_obj(bind_rows(user_info, user_state))
        }
      )
      output$hash_output <- shiny::renderText(encoded_txt())
    },
    envir = p
  )
}

hash_encoder_ui <- {
  shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please enter your identifying information, then click the button below to generate your hash",
    textInput("name", "What's your name?"),
    textInput("studentID", "What is your student ID?"),
    renderText({
      input$caption
    })
  )
}

solution_code_card <- function(code, title = "Solution") {
  tags$div(
    style = paste(
      "margin: 1rem 0 1.25rem;",
      "border: 1px solid #d7e3f3;",
      "border-left: 5px solid #2f6f9f;",
      "border-radius: 8px;",
      "background: #f8fbff;",
      "box-shadow: 0 2px 8px rgba(31, 45, 61, 0.08);",
      "overflow: hidden;"
    ),
    tags$div(
      style = paste(
        "display: flex;",
        "align-items: center;",
        "justify-content: space-between;",
        "gap: 0.75rem;",
        "padding: 0.7rem 0.9rem;",
        "background: #eef6ff;",
        "border-bottom: 1px solid #d7e3f3;"
      ),
      tags$strong(title, style = "color: #1f4e79; font-size: 1rem;"),
      tags$span(
        "released",
        style = paste(
          "padding: 0.15rem 0.5rem;",
          "border-radius: 999px;",
          "background: #ffffff;",
          "color: #3c6478;",
          "font-size: 0.78rem;",
          "font-weight: 600;"
        )
      )
    ),
    tags$pre(
      style = paste(
        "margin: 0;",
        "padding: 0.9rem 1rem;",
        "background: #ffffff;",
        "border: 0;",
        "white-space: pre-wrap;",
        "text-align: left;"
      ),
      tags$code(
        htmltools::HTML(htmltools::htmlEscape(code)),
        style = "display: block; color: #243447; font-size: 0.95rem; line-height: 1.45;"
      )
    )
  )
}

timed_solution <- function(output_id, code, title = "Solution",
                           reveal_time = my_reveal_time,
                           tz = "America/New_York",
                           check_every_ms = 1000) {
  p <- parent.frame()
  check_server_context(p)

  p$output[[output_id]] <- shiny::renderUI({
    shiny::invalidateLater(check_every_ms, p$session)
    release_at <- as.POSIXct(reveal_time, tz = tz)
    if (Sys.time() < release_at) {
      return(NULL)
    }
    solution_code_card(code, title)
  })
}
```

## Welcome

Hello, and welcome to **Bayesian Updating**!

This tutorial is preparation for **Lab 7**. In Lab 7 you will use
Bayes' rule, likelihoods, priors, and posterior distributions to explain
how evidence updates belief.

Our learning goals are to:

- use Bayes' rule with a small probability table;
- compute a likelihood for observed successes and failures;
- build a posterior distribution on a grid;
- compare weak and strong priors;
- describe Bayesian and frequentist conclusions in different language.

## Packages

```{r load-packages, exercise=TRUE}
library(dplyr)
library(ggplot2)
```

## Bayes' Rule with Counts

Suppose 1% of people have a condition. A test is positive for 95% of
people with the condition and 10% of people without it.

```{r diagnostic-counts, exercise=TRUE}
diagnostic_table <- tibble(
  group = c("condition", "no_condition"),
  people = c(100, 9900),
  positive_rate = c(0.95, 0.10)
) %>%
  mutate(
    positive = people * positive_rate,
    negative = people - positive
  )

diagnostic_table
```

```{r posterior-diagnostic, exercise=TRUE}
diagnostic_table %>%
  summarise(
    posterior_condition_given_positive =
      positive[group == "condition"] / sum(positive)
  )
```

## Your Turn 1

Change the false positive rate from 10% to 5%. What happens to the
posterior probability of having the condition after a positive test?

```{r ex-false-positive, exercise=TRUE}

```

```{r ex-false-positive-hint-1}
# Change positive_rate for the no_condition row from 0.10 to 0.05.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_false_positive_solution_code",
  "diagnostic_table <- tibble(\n  group = c(\"condition\", \"no_condition\"),\n  people = c(100, 9900),\n  positive_rate = c(0.95, 0.05)\n) %>%\n  mutate(\n    positive = people * positive_rate,\n    negative = people - positive\n  )\n\ndiagnostic_table %>%\n  summarise(\n    posterior_condition_given_positive =\n      positive[group == \"condition\"] / sum(positive)\n  )"
)
```

```{r ex-false-positive-solution-ui, echo=FALSE}
uiOutput("ex_false_positive_solution_code")
```

## A Grid Posterior

Now suppose a coin lands heads 8 times in 10 flips. We will estimate
the probability of heads, called `theta`, using a grid of possible
values.

```{r grid-prior, exercise=TRUE}
theta_grid <- seq(0, 1, length.out = 101)

coin_grid <- tibble(
  theta = theta_grid,
  prior = 1
)

coin_grid
```

```{r likelihood, exercise=TRUE}
heads <- 8
flips <- 10

coin_grid <- coin_grid %>%
  mutate(
    likelihood = dbinom(heads, size = flips, prob = theta),
    unscaled_posterior = prior * likelihood,
    posterior = unscaled_posterior / sum(unscaled_posterior)
  )

coin_grid
```

```{r posterior-plot, exercise=TRUE}
ggplot(coin_grid, aes(x = theta, y = posterior)) +
  geom_col(width = 0.01) +
  labs(x = "Possible value of theta", y = "Posterior probability")
```

## Posterior Summaries

```{r posterior-summary, exercise=TRUE}
coin_grid %>%
  summarise(
    posterior_mean = sum(theta * posterior),
    posterior_probability_theta_above_half = sum(posterior[theta > 0.5])
  )
```

## Your Turn 2

Rebuild the posterior for 14 heads in 20 flips. Compute the posterior
mean.

```{r ex-more-data, exercise=TRUE}
heads <- 14
flips <- 20
```

```{r ex-more-data-hint-1}
# Reuse the likelihood, unscaled posterior, and posterior calculations.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_more_data_solution_code",
  "heads <- 14\nflips <- 20\n\ncoin_grid <- tibble(theta = theta_grid, prior = 1) %>%\n  mutate(\n    likelihood = dbinom(heads, size = flips, prob = theta),\n    unscaled_posterior = prior * likelihood,\n    posterior = unscaled_posterior / sum(unscaled_posterior)\n  )\n\ncoin_grid %>%\n  summarise(posterior_mean = sum(theta * posterior))"
)
```

```{r ex-more-data-solution-ui, echo=FALSE}
uiOutput("ex_more_data_solution_code")
```

## Compare Priors

A stronger prior can pull the posterior toward the values the prior
favored, especially when the amount of data is small.

```{r compare-priors, exercise=TRUE}
prior_comparison <- tibble(
  theta = rep(theta_grid, times = 2),
  prior_type = rep(c("uniform", "skeptical"), each = length(theta_grid)),
  prior = c(
    rep(1, length(theta_grid)),
    dbeta(theta_grid, shape1 = 20, shape2 = 20)
  )
) %>%
  group_by(prior_type) %>%
  mutate(
    likelihood = dbinom(8, size = 10, prob = theta),
    posterior = prior * likelihood / sum(prior * likelihood)
  ) %>%
  ungroup()

ggplot(prior_comparison, aes(x = theta, y = posterior, color = prior_type)) +
  geom_line() +
  labs(x = "Possible value of theta", y = "Posterior probability")
```

## Interpretation Check

In one or two sentences, explain why two people can use the same data
but get different posterior distributions.

```{r ex-prior-interpretation, exercise=TRUE}

```

## Final Task

Use a uniform prior and data with 18 heads in 30 flips. Plot the
posterior and compute the posterior probability that `theta > 0.5`.
Then write one sentence interpreting the probability.

```{r ex-final-posterior, exercise=TRUE}

```

```{r ex-final-posterior-hint-1}
# Set heads <- 18 and flips <- 30, then reuse the grid posterior pattern.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_final_posterior_solution_code",
  "heads <- 18\nflips <- 30\n\nfinal_grid <- tibble(theta = theta_grid, prior = 1) %>%\n  mutate(\n    likelihood = dbinom(heads, size = flips, prob = theta),\n    posterior = prior * likelihood / sum(prior * likelihood)\n  )\n\nggplot(final_grid, aes(x = theta, y = posterior)) +\n  geom_col(width = 0.01) +\n  labs(x = \"Possible value of theta\", y = \"Posterior probability\")\n\nfinal_grid %>%\n  summarise(probability_theta_above_half = sum(posterior[theta > 0.5]))"
)
```

```{r ex-final-posterior-solution-ui, echo=FALSE}
uiOutput("ex_final_posterior_solution_code")
```

## Submit

```{r, echo=FALSE, context="server"}
encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(ui_before = hash_encoder_ui)
```
