---
title: "Tutorial 5: LLN, CLT, and Sampling Distributions"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(shiny)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
tutorial_options(exercise.eval = FALSE)

my_reveal_time <- "2026-06-20 15:00:00"

population <- tibble(value = rgamma(10000, shape = 2, rate = 0.5))

is_server_context <- function(.envir) {
  inherits(.envir$input, "reactivevalues") &
    inherits(.envir$output, "shinyoutput") &
    inherits(.envir$session, "ShinySession")
}

check_server_context <- function(.envir) {
  if (!is_server_context(.envir)) {
    calling_func <- deparse(sys.calls()[[sys.nframe() - 1]])
    err <- paste0("Function `", calling_func, "` must be called from an Rmd chunk where `context = \"server\"`")
    stop(err, call. = FALSE)
  }
}

encoder_logic <- function(strip_output = FALSE) {
  p <- parent.frame()
  check_server_context(p)
  assign("strip_output", strip_output, envir = p)
  local(
    {
      encoded_txt <- shiny::eventReactive(
        input$hash_generate,
        {
          state <- learnr:::get_tutorial_state()
          shiny::validate(shiny::need(length(state) > 0, "No progress yet."))
          shiny::validate(shiny::need(nchar(input$name) > 0, "No name entered."))
          shiny::validate(shiny::need(nchar(input$studentID) > 0, "Please enter your student ID"))
          user_state <- purrr::map_dfr(state, identity, .id = "label")
          user_state <- dplyr::group_by(user_state, label, type, correct)
          user_state <- dplyr::summarize(
            user_state,
            answer = list(answer),
            timestamp = dplyr::first(timestamp),
            .groups = "drop"
          )
          user_state <- dplyr::relocate(user_state, correct, .before = timestamp)
          user_info <- tibble(
            label = c("student_name", "student_id"),
            type = "identifier",
            answer = as.list(c(input$name, input$studentID)),
            timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "UTC")
          )
          learnrhash::encode_obj(bind_rows(user_info, user_state))
        }
      )
      output$hash_output <- shiny::renderText(encoded_txt())
    },
    envir = p
  )
}

hash_encoder_ui <- {
  shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please enter your identifying information, then click the button below to generate your hash",
    textInput("name", "What's your name?"),
    textInput("studentID", "What is your student ID?"),
    renderText({
      input$caption
    })
  )
}

solution_code_card <- function(code, title = "Solution") {
  tags$div(
    style = paste(
      "margin: 1rem 0 1.25rem;",
      "border: 1px solid #d7e3f3;",
      "border-left: 5px solid #2f6f9f;",
      "border-radius: 8px;",
      "background: #f8fbff;",
      "box-shadow: 0 2px 8px rgba(31, 45, 61, 0.08);",
      "overflow: hidden;"
    ),
    tags$div(
      style = paste(
        "display: flex;",
        "align-items: center;",
        "justify-content: space-between;",
        "gap: 0.75rem;",
        "padding: 0.7rem 0.9rem;",
        "background: #eef6ff;",
        "border-bottom: 1px solid #d7e3f3;"
      ),
      tags$strong(title, style = "color: #1f4e79; font-size: 1rem;"),
      tags$span(
        "released",
        style = paste(
          "padding: 0.15rem 0.5rem;",
          "border-radius: 999px;",
          "background: #ffffff;",
          "color: #3c6478;",
          "font-size: 0.78rem;",
          "font-weight: 600;"
        )
      )
    ),
    tags$pre(
      style = paste(
        "margin: 0;",
        "padding: 0.9rem 1rem;",
        "background: #ffffff;",
        "border: 0;",
        "white-space: pre-wrap;",
        "text-align: left;"
      ),
      tags$code(
        htmltools::HTML(htmltools::htmlEscape(code)),
        style = "display: block; color: #243447; font-size: 0.95rem; line-height: 1.45;"
      )
    )
  )
}

timed_solution <- function(output_id, code, title = "Solution",
                           reveal_time = my_reveal_time,
                           tz = "America/New_York",
                           check_every_ms = 1000) {
  p <- parent.frame()
  check_server_context(p)

  p$output[[output_id]] <- shiny::renderUI({
    shiny::invalidateLater(check_every_ms, p$session)
    release_at <- as.POSIXct(reveal_time, tz = tz)
    if (Sys.time() < release_at) {
      return(NULL)
    }
    solution_code_card(code, title)
  })
}
```

## Welcome

Hello, and welcome to **LLN, CLT, and Sampling Distributions**!

This tutorial is preparation for **Lab 5**. In Lab 5 you will simulate
sampling distributions from several different populations and explain
how the Law of Large Numbers and Central Limit Theorem show up in the
results.

Our learning goals are to:

- generate values from a population distribution;
- draw one sample and compute a sample mean;
- repeat sampling to build a sampling distribution;
- compare the effect of sample size `n` and repetitions `K`;
- connect simulated spread to standard error.

## Packages

```{r load-packages, exercise=TRUE}
library(dplyr)
library(ggplot2)
```

## A Skewed Population

We will start with a right-skewed population. The population mean is the
long-run target for sample means.

```{r create-population, exercise=TRUE}
set.seed(213)

population <- tibble(value = rgamma(10000, shape = 2, rate = 0.5))

population %>%
  summarise(
    population_mean = mean(value),
    population_sd = sd(value)
  )
```

```{r population-plot, exercise=TRUE}
ggplot(population, aes(x = value)) +
  geom_histogram(bins = 40, color = "white") +
  labs(x = "Population value", y = "Count")
```

## Take One Sample

```{r one-sample, exercise=TRUE}
set.seed(213)

one_sample <- population %>%
  slice_sample(n = 30)

one_sample %>%
  summarise(sample_mean = mean(value))
```

## Your Turn 1

Take a sample of 10 observations and compute its sample mean.

```{r ex-one-small-sample, exercise=TRUE}
set.seed(213)
```

```{r ex-one-small-sample-hint-1}
# Use slice_sample(n = 10), then summarise(sample_mean = mean(value)).
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_one_small_sample_solution_code",
  "set.seed(213)\n\nsmall_sample <- population %>%\n  slice_sample(n = 10)\n\nsmall_sample %>%\n  summarise(sample_mean = mean(value))"
)
```

```{r ex-one-small-sample-solution-ui, echo=FALSE}
uiOutput("ex_one_small_sample_solution_code")
```

## Build a Sampling Distribution

Now we repeat the sampling process many times. Each repetition produces
one sample mean.

```{r repeated-samples, exercise=TRUE}
set.seed(213)

sampling_distribution <- replicate(
  n = 1000,
  expr = {
    population %>%
      slice_sample(n = 30) %>%
      summarise(sample_mean = mean(value)) %>%
      pull(sample_mean)
  }
)

sampling_distribution <- tibble(sample_mean = sampling_distribution)

head(sampling_distribution)
```

```{r sampling-distribution-plot, exercise=TRUE}
ggplot(sampling_distribution, aes(x = sample_mean)) +
  geom_histogram(bins = 35, color = "white") +
  labs(x = "Sample mean", y = "Number of samples")
```

## Your Turn 2

Repeat the sampling process with samples of size 100 instead of 30.
Then plot the sampling distribution.

```{r ex-large-sampling-distribution, exercise=TRUE}
set.seed(213)
```

```{r ex-large-sampling-distribution-hint-1}
# Copy the replicate() pattern above and change n = 30 to n = 100.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_large_sampling_distribution_solution_code",
  "set.seed(213)\n\nsampling_distribution_100 <- replicate(\n  n = 1000,\n  expr = {\n    population %>%\n      slice_sample(n = 100) %>%\n      summarise(sample_mean = mean(value)) %>%\n      pull(sample_mean)\n  }\n)\n\nsampling_distribution_100 <- tibble(sample_mean = sampling_distribution_100)\n\nggplot(sampling_distribution_100, aes(x = sample_mean)) +\n  geom_histogram(bins = 35, color = \"white\") +\n  labs(x = \"Sample mean\", y = \"Number of samples\")"
)
```

```{r ex-large-sampling-distribution-solution-ui, echo=FALSE}
uiOutput("ex_large_sampling_distribution_solution_code")
```

## Standard Error

For a sample mean, the standard error is:

$$
SE = \frac{\sigma}{\sqrt{n}}
$$

In practice we often estimate `sigma` with the population or sample
standard deviation available in the simulation.

```{r standard-error, exercise=TRUE}
population_sd <- sd(population$value)
n <- 30

standard_error <- population_sd / sqrt(n)
standard_error
```

```{r simulated-spread, exercise=TRUE}
sampling_distribution %>%
  summarise(simulated_sd = sd(sample_mean))
```

## Your Turn 3

Write a function named `simulate_means()` that takes a sample size `n`
and number of repetitions `K`, then returns a tibble of sample means.

```{r ex-simulate-means-function, exercise=TRUE}

```

```{r ex-simulate-means-function-hint-1}
# Use replicate(n = K, expr = {...}) inside the function.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_simulate_means_function_solution_code",
  "simulate_means <- function(n, K) {\n  means <- replicate(\n    n = K,\n    expr = {\n      population %>%\n        slice_sample(n = n) %>%\n        summarise(sample_mean = mean(value)) %>%\n        pull(sample_mean)\n    }\n  )\n\n  tibble(sample_mean = means)\n}\n\nsimulate_means(n = 30, K = 5)"
)
```

```{r ex-simulate-means-function-solution-ui, echo=FALSE}
uiOutput("ex_simulate_means_function_solution_code")
```

## Interpretation Check

In one or two sentences, explain why the sampling distribution with
larger `n` is narrower than the sampling distribution with smaller `n`.

```{r ex-lln-clt-interpretation, exercise=TRUE}

```

## Final Task

Compare sample sizes 10, 30, and 100 using `K = 1000` repetitions.
Create a table with the simulated center and spread for each sample
size, then write two sentences connecting the pattern to the Central
Limit Theorem.

```{r ex-final-sample-size, exercise=TRUE}
sample_sizes <- c(10, 30, 100)
```

```{r ex-final-sample-size-hint-1}
# Use simulate_means() for each sample size, then summarise().
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_final_sample_size_solution_code",
  "simulate_means <- function(n, K) {\n  means <- replicate(\n    n = K,\n    expr = {\n      population %>%\n        slice_sample(n = n) %>%\n        summarise(sample_mean = mean(value)) %>%\n        pull(sample_mean)\n    }\n  )\n\n  tibble(sample_mean = means)\n}\n\nsample_sizes <- c(10, 30, 100)\n\nsummary_table <- purrr::map_dfr(\n  sample_sizes,\n  ~ simulate_means(n = .x, K = 1000) %>%\n    summarise(\n      n = .x,\n      center = mean(sample_mean),\n      spread = sd(sample_mean)\n    )\n)\n\nsummary_table"
)
```

```{r ex-final-sample-size-solution-ui, echo=FALSE}
uiOutput("ex_final_sample_size_solution_code")
```

## Submit

```{r, echo=FALSE, context="server"}
encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(ui_before = hash_encoder_ui)
```
