---
title: "Inference: 5 - Simulation, Sampling Distributions, and Inference"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(shiny)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
tutorial_options(exercise.eval = FALSE)

my_reveal_time <- "2026-06-20 15:00:00"

population <- tibble(
  student = 1:10000,
  supports_policy = rbinom(10000, size = 1, prob = 0.58)
)

is_server_context <- function(.envir) {
  inherits(.envir$input, "reactivevalues") &
    inherits(.envir$output, "shinyoutput") &
    inherits(.envir$session, "ShinySession")
}

check_server_context <- function(.envir) {
  if (!is_server_context(.envir)) {
    calling_func <- deparse(sys.calls()[[sys.nframe() - 1]])
    err <- paste0("Function `", calling_func, "` must be called from an Rmd chunk where `context = \"server\"`")
    stop(err, call. = FALSE)
  }
}

encoder_logic <- function(strip_output = FALSE) {
  p <- parent.frame()
  check_server_context(p)
  assign("strip_output", strip_output, envir = p)
  local(
    {
      encoded_txt <- shiny::eventReactive(
        input$hash_generate,
        {
          state <- learnr:::get_tutorial_state()
          shiny::validate(shiny::need(length(state) > 0, "No progress yet."))
          shiny::validate(shiny::need(nchar(input$name) > 0, "No name entered."))
          shiny::validate(shiny::need(nchar(input$studentID) > 0, "Please enter your student ID"))
          user_state <- purrr::map_dfr(state, identity, .id = "label")
          user_state <- dplyr::group_by(user_state, label, type, correct)
          user_state <- dplyr::summarize(
            user_state,
            answer = list(answer),
            timestamp = dplyr::first(timestamp),
            .groups = "drop"
          )
          user_state <- dplyr::relocate(user_state, correct, .before = timestamp)
          user_info <- tibble(
            label = c("student_name", "student_id"),
            type = "identifier",
            answer = as.list(c(input$name, input$studentID)),
            timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "UTC")
          )
          learnrhash::encode_obj(bind_rows(user_info, user_state))
        }
      )
      output$hash_output <- shiny::renderText(encoded_txt())
    },
    envir = p
  )
}

hash_encoder_ui <- {
  shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please enter your identifying information, then click the button below to generate your hash",
    textInput("name", "What's your name?"),
    textInput("studentID", "What is your student ID?"),
    renderText({
      input$caption
    })
  )
}

solution_code_card <- function(code, title = "Solution") {
  tags$div(
    style = paste(
      "margin: 1rem 0 1.25rem;",
      "border: 1px solid #d7e3f3;",
      "border-left: 5px solid #2f6f9f;",
      "border-radius: 8px;",
      "background: #f8fbff;",
      "box-shadow: 0 2px 8px rgba(31, 45, 61, 0.08);",
      "overflow: hidden;"
    ),
    tags$div(
      style = paste(
        "display: flex;",
        "align-items: center;",
        "justify-content: space-between;",
        "gap: 0.75rem;",
        "padding: 0.7rem 0.9rem;",
        "background: #eef6ff;",
        "border-bottom: 1px solid #d7e3f3;"
      ),
      tags$strong(title, style = "color: #1f4e79; font-size: 1rem;"),
      tags$span(
        "released",
        style = paste(
          "padding: 0.15rem 0.5rem;",
          "border-radius: 999px;",
          "background: #ffffff;",
          "color: #3c6478;",
          "font-size: 0.78rem;",
          "font-weight: 600;"
        )
      )
    ),
    tags$pre(
      style = paste(
        "margin: 0;",
        "padding: 0.9rem 1rem;",
        "background: #ffffff;",
        "border: 0;",
        "white-space: pre-wrap;",
        "text-align: left;"
      ),
      tags$code(
        htmltools::HTML(htmltools::htmlEscape(code)),
        style = "display: block; color: #243447; font-size: 0.95rem; line-height: 1.45;"
      )
    )
  )
}

timed_solution <- function(output_id, code, title = "Solution",
                           reveal_time = my_reveal_time,
                           tz = "America/New_York",
                           check_every_ms = 1000) {
  p <- parent.frame()
  check_server_context(p)

  p$output[[output_id]] <- shiny::renderUI({
    shiny::invalidateLater(check_every_ms, p$session)
    release_at <- as.POSIXct(reveal_time, tz = tz)
    if (Sys.time() < release_at) {
      return(NULL)
    }
    solution_code_card(code, title)
  })
}
```

## Welcome

Hello, and welcome to **Simulation, Sampling Distributions, and
Inference**!

In this tutorial we will repeatedly sample from a known population. The
goal is to see that sample statistics vary from sample to sample, and
that confidence intervals and p-values are built from this variability.

Our learning goals are to:

- compute a sample proportion;
- build a sampling distribution using repeated samples;
- compare sampling variability for small and large samples;
- construct and interpret a confidence interval for one proportion;
- compute and interpret a p-value for one proportion.

## Packages

```{r load-packages, exercise=TRUE}
library(dplyr)
library(ggplot2)
```

## A Simulated Population

Imagine a campus survey where each student either supports a proposed
policy or does not support it. The variable `supports_policy` is 1 for
support and 0 for no support.

```{r create-population, exercise=TRUE}
set.seed(213)

population <- tibble(
  student = 1:10000,
  supports_policy = rbinom(10000, size = 1, prob = 0.58)
)

population %>%
  summarise(parameter = mean(supports_policy))
```

## Take One Sample

```{r one-sample, exercise=TRUE}
set.seed(213)

one_sample <- population %>%
  slice_sample(n = 80)

one_sample %>%
  summarise(p_hat = mean(supports_policy))
```

## Your Turn 1

Take a sample of 40 students and compute the sample proportion who
support the policy.

```{r ex-one-small-sample, exercise=TRUE}
set.seed(213)
```

```{r ex-one-small-sample-hint-1}
# Use slice_sample(n = 40), then summarise().
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_one_small_sample_solution_code",
  "set.seed(213)\n\nsmall_sample <- population %>%\n  slice_sample(n = 40)\n\nsmall_sample %>%\n  summarise(p_hat = mean(supports_policy))"
)
```

```{r ex-one-small-sample-solution-ui, echo=FALSE}
uiOutput("ex_one_small_sample_solution_code")
```

## Repeat the Sampling Process

Now we will take many samples. Each sample gives one value of
`p_hat`. Together, those values form a sampling distribution.

```{r repeated-samples, exercise=TRUE}
set.seed(213)

sampling_distribution <- replicate(
  n = 1000,
  expr = {
    population %>%
      slice_sample(n = 80) %>%
      summarise(p_hat = mean(supports_policy)) %>%
      pull(p_hat)
  }
)

sampling_distribution <- tibble(p_hat = sampling_distribution)

head(sampling_distribution)
```

## Plot the Sampling Distribution

```{r sampling-distribution-plot, exercise=TRUE}
ggplot(sampling_distribution, aes(x = p_hat)) +
  geom_histogram(binwidth = 0.025, color = "white") +
  labs(x = "Sample proportion", y = "Number of samples")
```

## Your Turn 2

Repeat the sampling process with samples of size 200 instead of 80.
Then plot the sampling distribution.

```{r ex-large-sampling-distribution, exercise=TRUE}
set.seed(213)
```

```{r ex-large-sampling-distribution-hint-1}
# Copy the replicate() pattern above and change n = 80 to n = 200.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_large_sampling_distribution_solution_code",
  "set.seed(213)\n\nsampling_distribution_200 <- replicate(\n  n = 1000,\n  expr = {\n    population %>%\n      slice_sample(n = 200) %>%\n      summarise(p_hat = mean(supports_policy)) %>%\n      pull(p_hat)\n  }\n)\n\nsampling_distribution_200 <- tibble(p_hat = sampling_distribution_200)\n\nggplot(sampling_distribution_200, aes(x = p_hat)) +\n  geom_histogram(binwidth = 0.02, color = \"white\") +\n  labs(x = \"Sample proportion\", y = \"Number of samples\")"
)
```

```{r ex-large-sampling-distribution-solution-ui, echo=FALSE}
uiOutput("ex_large_sampling_distribution_solution_code")
```

## Standard Error

For one proportion, the standard error is:

$$
SE = \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}
$$

```{r standard-error, exercise=TRUE}
p_hat <- mean(one_sample$supports_policy)
n <- nrow(one_sample)

standard_error <- sqrt(p_hat * (1 - p_hat) / n)
standard_error
```

## Confidence Interval

A rough 95% confidence interval for one proportion is:

$$
\hat{p} \pm 1.96 \times SE
$$

```{r confidence-interval, exercise=TRUE}
lower <- p_hat - 1.96 * standard_error
upper <- p_hat + 1.96 * standard_error

c(lower, upper)
```

## Your Turn 3

Write a function named `prop_ci()` that takes a sample proportion
`p_hat` and a sample size `n`, then returns the lower and upper bounds
of a 95% confidence interval.

```{r ex-prop-ci-function, exercise=TRUE}

```

```{r ex-prop-ci-function-hint-1}
# Return c(lower, upper).
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_prop_ci_function_solution_code",
  "prop_ci <- function(p_hat, n) {\n  se <- sqrt(p_hat * (1 - p_hat) / n)\n  lower <- p_hat - 1.96 * se\n  upper <- p_hat + 1.96 * se\n  c(lower, upper)\n}\n\nprop_ci(p_hat = 0.60, n = 80)"
)
```

```{r ex-prop-ci-function-solution-ui, echo=FALSE}
uiOutput("ex_prop_ci_function_solution_code")
```

## Hypothesis Test

Suppose the null hypothesis is that exactly half of students support the
policy.

$$
H_0: p = 0.50
$$

We can approximate a two-sided p-value using a z statistic.

```{r p-value-example, exercise=TRUE}
null_p <- 0.50
se_null <- sqrt(null_p * (1 - null_p) / n)
z <- (p_hat - null_p) / se_null
p_value <- 2 * (1 - pnorm(abs(z)))

p_value
```

## Interpretation Check

In one or two sentences, interpret the confidence interval from this
tutorial in the context of campus policy support.

```{r ex-ci-interpretation, exercise=TRUE}

```

## Final Task

Investigate how sample size changes confidence interval width. Use
`p_hat = 0.58` with sample sizes 40, 80, 160, and 320. Then write two
sentences explaining the pattern.

```{r ex-final-ci-width, exercise=TRUE}
sample_sizes <- c(40, 80, 160, 320)
```

```{r ex-final-ci-width-hint-1}
# Use your prop_ci() function or repeat the formula.
# Width is upper bound minus lower bound.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_final_ci_width_solution_code",
  "prop_ci <- function(p_hat, n) {\n  se <- sqrt(p_hat * (1 - p_hat) / n)\n  lower <- p_hat - 1.96 * se\n  upper <- p_hat + 1.96 * se\n  c(lower, upper)\n}\n\nsample_sizes <- c(40, 80, 160, 320)\n\nci_widths <- tibble(n = sample_sizes) %>%\n  rowwise() %>%\n  mutate(width = diff(prop_ci(0.58, n))) %>%\n  ungroup()\n\nci_widths"
)
```

```{r ex-final-ci-width-solution-ui, echo=FALSE}
uiOutput("ex_final_ci_width_solution_code")
```

## Submit

```{r, echo=FALSE, context="server"}
encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(ui_before = hash_encoder_ui)
```
