---
title: "Tutorial 6: Confidence Intervals and Hypothesis Tests"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(shiny)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
tutorial_options(exercise.eval = FALSE)

my_reveal_time <- "2026-06-20 15:00:00"

population <- tibble(
  person = 1:10000,
  supports_policy = rbinom(10000, size = 1, prob = 0.58)
)

prop_ci <- function(p_hat, n) {
  se <- sqrt(p_hat * (1 - p_hat) / n)
  lower <- p_hat - 1.96 * se
  upper <- p_hat + 1.96 * se
  c(lower, upper)
}

is_server_context <- function(.envir) {
  inherits(.envir$input, "reactivevalues") &
    inherits(.envir$output, "shinyoutput") &
    inherits(.envir$session, "ShinySession")
}

check_server_context <- function(.envir) {
  if (!is_server_context(.envir)) {
    calling_func <- deparse(sys.calls()[[sys.nframe() - 1]])
    err <- paste0("Function `", calling_func, "` must be called from an Rmd chunk where `context = \"server\"`")
    stop(err, call. = FALSE)
  }
}

encoder_logic <- function(strip_output = FALSE) {
  p <- parent.frame()
  check_server_context(p)
  assign("strip_output", strip_output, envir = p)
  local(
    {
      encoded_txt <- shiny::eventReactive(
        input$hash_generate,
        {
          state <- learnr:::get_tutorial_state()
          shiny::validate(shiny::need(length(state) > 0, "No progress yet."))
          shiny::validate(shiny::need(nchar(input$name) > 0, "No name entered."))
          shiny::validate(shiny::need(nchar(input$studentID) > 0, "Please enter your student ID"))
          user_state <- purrr::map_dfr(state, identity, .id = "label")
          user_state <- dplyr::group_by(user_state, label, type, correct)
          user_state <- dplyr::summarize(
            user_state,
            answer = list(answer),
            timestamp = dplyr::first(timestamp),
            .groups = "drop"
          )
          user_state <- dplyr::relocate(user_state, correct, .before = timestamp)
          user_info <- tibble(
            label = c("student_name", "student_id"),
            type = "identifier",
            answer = as.list(c(input$name, input$studentID)),
            timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "UTC")
          )
          learnrhash::encode_obj(bind_rows(user_info, user_state))
        }
      )
      output$hash_output <- shiny::renderText(encoded_txt())
    },
    envir = p
  )
}

hash_encoder_ui <- {
  shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please enter your identifying information, then click the button below to generate your hash",
    textInput("name", "What's your name?"),
    textInput("studentID", "What is your student ID?"),
    renderText({
      input$caption
    })
  )
}

solution_code_card <- function(code, title = "Solution") {
  tags$div(
    style = paste(
      "margin: 1rem 0 1.25rem;",
      "border: 1px solid #d7e3f3;",
      "border-left: 5px solid #2f6f9f;",
      "border-radius: 8px;",
      "background: #f8fbff;",
      "box-shadow: 0 2px 8px rgba(31, 45, 61, 0.08);",
      "overflow: hidden;"
    ),
    tags$div(
      style = paste(
        "display: flex;",
        "align-items: center;",
        "justify-content: space-between;",
        "gap: 0.75rem;",
        "padding: 0.7rem 0.9rem;",
        "background: #eef6ff;",
        "border-bottom: 1px solid #d7e3f3;"
      ),
      tags$strong(title, style = "color: #1f4e79; font-size: 1rem;"),
      tags$span(
        "released",
        style = paste(
          "padding: 0.15rem 0.5rem;",
          "border-radius: 999px;",
          "background: #ffffff;",
          "color: #3c6478;",
          "font-size: 0.78rem;",
          "font-weight: 600;"
        )
      )
    ),
    tags$pre(
      style = paste(
        "margin: 0;",
        "padding: 0.9rem 1rem;",
        "background: #ffffff;",
        "border: 0;",
        "white-space: pre-wrap;",
        "text-align: left;"
      ),
      tags$code(
        htmltools::HTML(htmltools::htmlEscape(code)),
        style = "display: block; color: #243447; font-size: 0.95rem; line-height: 1.45;"
      )
    )
  )
}

timed_solution <- function(output_id, code, title = "Solution",
                           reveal_time = my_reveal_time,
                           tz = "America/New_York",
                           check_every_ms = 1000) {
  p <- parent.frame()
  check_server_context(p)

  p$output[[output_id]] <- shiny::renderUI({
    shiny::invalidateLater(check_every_ms, p$session)
    release_at <- as.POSIXct(reveal_time, tz = tz)
    if (Sys.time() < release_at) {
      return(NULL)
    }
    solution_code_card(code, title)
  })
}
```

## Welcome

Hello, and welcome to **Confidence Intervals and Hypothesis Tests**!

This tutorial is preparation for **Lab 6**. In Lab 6 you will compare
confidence intervals and hypothesis tests, simulate confidence interval
coverage, and build a chi-square test of independence.

Our learning goals are to:

- compute a sample proportion;
- construct and interpret a confidence interval;
- state and test a null hypothesis for one proportion;
- simulate confidence interval coverage;
- organize categorical data for a chi-square test.

## Packages

```{r load-packages, exercise=TRUE}
library(dplyr)
library(ggplot2)
```

## A Simulated Population

Imagine a campus survey where each person either supports a proposed
policy or does not support it. The variable `supports_policy` is 1 for
support and 0 for no support.

```{r create-population, exercise=TRUE}
set.seed(213)

population <- tibble(
  person = 1:10000,
  supports_policy = rbinom(10000, size = 1, prob = 0.58)
)

population %>%
  summarise(parameter = mean(supports_policy))
```

## Take One Sample

```{r one-sample, exercise=TRUE}
set.seed(213)

one_sample <- population %>%
  slice_sample(n = 80)

one_sample %>%
  summarise(p_hat = mean(supports_policy))
```

## Confidence Interval

For one proportion, the standard error is:

$$
SE = \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}
$$

```{r confidence-interval, exercise=TRUE}
p_hat <- mean(one_sample$supports_policy)
n <- nrow(one_sample)
standard_error <- sqrt(p_hat * (1 - p_hat) / n)

lower <- p_hat - 1.96 * standard_error
upper <- p_hat + 1.96 * standard_error

c(lower, upper)
```

## Your Turn 1

Write a function named `prop_ci()` that takes a sample proportion
`p_hat` and a sample size `n`, then returns the lower and upper bounds
of a 95% confidence interval.

```{r ex-prop-ci-function, exercise=TRUE}

```

```{r ex-prop-ci-function-hint-1}
# Return c(lower, upper).
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_prop_ci_function_solution_code",
  "prop_ci <- function(p_hat, n) {\n  se <- sqrt(p_hat * (1 - p_hat) / n)\n  lower <- p_hat - 1.96 * se\n  upper <- p_hat + 1.96 * se\n  c(lower, upper)\n}\n\nprop_ci(p_hat = 0.60, n = 80)"
)
```

```{r ex-prop-ci-function-solution-ui, echo=FALSE}
uiOutput("ex_prop_ci_function_solution_code")
```

## Hypothesis Test

Suppose the null hypothesis is that exactly half of people support the
policy.

$$
H_0: p = 0.50
$$

We can approximate a two-sided p-value using a z statistic.

```{r p-value-example, exercise=TRUE}
null_p <- 0.50
se_null <- sqrt(null_p * (1 - null_p) / n)
z <- (p_hat - null_p) / se_null
p_value <- 2 * (1 - pnorm(abs(z)))

p_value
```

## Your Turn 2

Repeat the one-proportion test using `null_p <- 0.60`. Is the p-value
larger or smaller than the p-value for `null_p <- 0.50`?

```{r ex-new-null, exercise=TRUE}
null_p <- 0.60
```

```{r ex-new-null-hint-1}
# Recompute se_null, z, and p_value using the new null value.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_new_null_solution_code",
  "null_p <- 0.60\nse_null <- sqrt(null_p * (1 - null_p) / n)\nz <- (p_hat - null_p) / se_null\np_value <- 2 * (1 - pnorm(abs(z)))\n\np_value"
)
```

```{r ex-new-null-solution-ui, echo=FALSE}
uiOutput("ex_new_null_solution_code")
```

## Simulate Coverage

Coverage asks how often an interval-building method captures the true
parameter across many repeated samples.

```{r simulate-coverage, exercise=TRUE}
set.seed(213)

capture_results <- replicate(
  n = 1000,
  expr = {
    sample_data <- population %>%
      slice_sample(n = 80)
    phat <- mean(sample_data$supports_policy)
    ci <- prop_ci(phat, n = 80)
    ci[1] <= 0.58 & 0.58 <= ci[2]
  }
)

mean(capture_results)
```

## Chi-Square Setup

For a chi-square test of independence, start with counts in a two-way
table.

```{r two-way-table, exercise=TRUE}
titanic_counts <- as.data.frame(Titanic) %>%
  filter(Age == "Adult") %>%
  group_by(Class, Survived) %>%
  summarise(count = sum(Freq), .groups = "drop")

titanic_counts
```

```{r chi-square-table, exercise=TRUE}
observed_table <- xtabs(count ~ Class + Survived, data = titanic_counts)
observed_table

chisq.test(observed_table)
```

## Interpretation Check

In one or two sentences, explain how a confidence interval answers a
different question from a hypothesis test.

```{r ex-ci-test-interpretation, exercise=TRUE}

```

## Final Task

Use `Titanic` to test whether passenger `Sex` and `Survived` appear
independent among adults. Write one sentence interpreting the p-value in
context.

```{r ex-final-chi-square, exercise=TRUE}

```

```{r ex-final-chi-square-hint-1}
# Filter Age == "Adult", aggregate counts by Sex and Survived, then use xtabs().
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_final_chi_square_solution_code",
  "sex_survival_counts <- as.data.frame(Titanic) %>%\n  filter(Age == \"Adult\") %>%\n  group_by(Sex, Survived) %>%\n  summarise(count = sum(Freq), .groups = \"drop\")\n\nsex_survival_table <- xtabs(count ~ Sex + Survived, data = sex_survival_counts)\n\nchisq.test(sex_survival_table)"
)
```

```{r ex-final-chi-square-solution-ui, echo=FALSE}
uiOutput("ex_final_chi_square_solution_code")
```

## Submit

```{r, echo=FALSE, context="server"}
encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(ui_before = hash_encoder_ui)
```
