---
title: "Modeling: 6 - Linear Modeling Workflow"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(shiny)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
tutorial_options(exercise.eval = FALSE)

my_reveal_time <- "2026-06-20 15:00:00"

air_model_data <- as_tibble(airquality) %>%
  drop_na(Ozone, Temp, Wind, Solar.R)

is_server_context <- function(.envir) {
  inherits(.envir$input, "reactivevalues") &
    inherits(.envir$output, "shinyoutput") &
    inherits(.envir$session, "ShinySession")
}

check_server_context <- function(.envir) {
  if (!is_server_context(.envir)) {
    calling_func <- deparse(sys.calls()[[sys.nframe() - 1]])
    err <- paste0("Function `", calling_func, "` must be called from an Rmd chunk where `context = \"server\"`")
    stop(err, call. = FALSE)
  }
}

encoder_logic <- function(strip_output = FALSE) {
  p <- parent.frame()
  check_server_context(p)
  assign("strip_output", strip_output, envir = p)
  local(
    {
      encoded_txt <- shiny::eventReactive(
        input$hash_generate,
        {
          state <- learnr:::get_tutorial_state()
          shiny::validate(shiny::need(length(state) > 0, "No progress yet."))
          shiny::validate(shiny::need(nchar(input$name) > 0, "No name entered."))
          shiny::validate(shiny::need(nchar(input$studentID) > 0, "Please enter your student ID"))
          user_state <- purrr::map_dfr(state, identity, .id = "label")
          user_state <- dplyr::group_by(user_state, label, type, correct)
          user_state <- dplyr::summarize(
            user_state,
            answer = list(answer),
            timestamp = dplyr::first(timestamp),
            .groups = "drop"
          )
          user_state <- dplyr::relocate(user_state, correct, .before = timestamp)
          user_info <- tibble(
            label = c("student_name", "student_id"),
            type = "identifier",
            answer = as.list(c(input$name, input$studentID)),
            timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "UTC")
          )
          learnrhash::encode_obj(bind_rows(user_info, user_state))
        }
      )
      output$hash_output <- shiny::renderText(encoded_txt())
    },
    envir = p
  )
}

hash_encoder_ui <- {
  shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please enter your identifying information, then click the button below to generate your hash",
    textInput("name", "What's your name?"),
    textInput("studentID", "What is your student ID?"),
    renderText({
      input$caption
    })
  )
}

solution_code_card <- function(code, title = "Solution") {
  tags$div(
    style = paste(
      "margin: 1rem 0 1.25rem;",
      "border: 1px solid #d7e3f3;",
      "border-left: 5px solid #2f6f9f;",
      "border-radius: 8px;",
      "background: #f8fbff;",
      "box-shadow: 0 2px 8px rgba(31, 45, 61, 0.08);",
      "overflow: hidden;"
    ),
    tags$div(
      style = paste(
        "display: flex;",
        "align-items: center;",
        "justify-content: space-between;",
        "gap: 0.75rem;",
        "padding: 0.7rem 0.9rem;",
        "background: #eef6ff;",
        "border-bottom: 1px solid #d7e3f3;"
      ),
      tags$strong(title, style = "color: #1f4e79; font-size: 1rem;"),
      tags$span(
        "released",
        style = paste(
          "padding: 0.15rem 0.5rem;",
          "border-radius: 999px;",
          "background: #ffffff;",
          "color: #3c6478;",
          "font-size: 0.78rem;",
          "font-weight: 600;"
        )
      )
    ),
    tags$pre(
      style = paste(
        "margin: 0;",
        "padding: 0.9rem 1rem;",
        "background: #ffffff;",
        "border: 0;",
        "white-space: pre-wrap;",
        "text-align: left;"
      ),
      tags$code(
        htmltools::HTML(htmltools::htmlEscape(code)),
        style = "display: block; color: #243447; font-size: 0.95rem; line-height: 1.45;"
      )
    )
  )
}

timed_solution <- function(output_id, code, title = "Solution",
                           reveal_time = my_reveal_time,
                           tz = "America/New_York",
                           check_every_ms = 1000) {
  p <- parent.frame()
  check_server_context(p)

  p$output[[output_id]] <- shiny::renderUI({
    shiny::invalidateLater(check_every_ms, p$session)
    release_at <- as.POSIXct(reveal_time, tz = tz)
    if (Sys.time() < release_at) {
      return(NULL)
    }
    solution_code_card(code, title)
  })
}
```

## Welcome

Hello, and welcome to **Linear Modeling Workflow**!

In this tutorial we will fit and interpret a simple linear model. The
coding goal is to use `lm()` and supporting plots. The statistical goal
is to explain what a model says in context and what it does not prove.

Our learning goals are to:

- make an exploratory scatterplot before modeling;
- fit a simple linear model with `lm()`;
- interpret the intercept and slope in context;
- add a fitted line to a scatterplot;
- inspect residuals and describe model limitations.

## Packages

```{r load-packages, exercise=TRUE}
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Prepare the Data

We will use the built-in `airquality` data. To keep the modeling
examples simple, we will remove rows missing `Ozone`, `Temp`, `Wind`, or
`Solar.R`.

```{r prepare-data, exercise=TRUE}
air_model_data <- as_tibble(airquality) %>%
  drop_na(Ozone, Temp, Wind, Solar.R)

glimpse(air_model_data)
```

## Explore Before Modeling

Before fitting a model, plot the response and explanatory variable. We
will model `Ozone` using `Temp`.

```{r exploratory-scatterplot, exercise=TRUE}
ggplot(air_model_data, aes(x = Temp, y = Ozone)) +
  geom_point() +
  labs(x = "Temperature", y = "Ozone")
```

## Fit a Linear Model

The formula `Ozone ~ Temp` means:

```         
response variable ~ explanatory variable
```

```{r fit-model, exercise=TRUE}
ozone_temp_model <- lm(Ozone ~ Temp, data = air_model_data)

summary(ozone_temp_model)
```

## Model Coefficients

```{r model-coefficients, exercise=TRUE}
coef(ozone_temp_model)
```

The intercept is the predicted ozone level when temperature is 0 degrees
Fahrenheit. The slope is the predicted change in ozone level for a
1-degree increase in temperature.

## Your Turn 1

Fit a model predicting `Ozone` from `Wind`. Save it as
`ozone_wind_model`, then print the model summary.

```{r ex-fit-wind-model, exercise=TRUE}

```

```{r ex-fit-wind-model-hint-1}
# Use lm(Ozone ~ Wind, data = air_model_data).
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_fit_wind_model_solution_code",
  "ozone_wind_model <- lm(Ozone ~ Wind, data = air_model_data)\n\nsummary(ozone_wind_model)"
)
```

```{r ex-fit-wind-model-solution-ui, echo=FALSE}
uiOutput("ex_fit_wind_model_solution_code")
```

## Add a Fitted Line

```{r fitted-line, exercise=TRUE}
ggplot(air_model_data, aes(x = Temp, y = Ozone)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Temperature", y = "Ozone")
```

## Predictions

Use `predict()` with a small data frame containing the explanatory
variable values where we want predictions.

```{r model-prediction, exercise=TRUE}
new_days <- tibble(Temp = c(70, 80, 90))

predict(ozone_temp_model, newdata = new_days)
```

## Your Turn 2

Use `ozone_wind_model` to predict ozone when wind speed is 5, 10, and
15 miles per hour.

```{r ex-wind-predictions, exercise=TRUE}

```

```{r ex-wind-predictions-hint-1}
# Your new data frame needs a column named Wind.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_wind_predictions_solution_code",
  "ozone_wind_model <- lm(Ozone ~ Wind, data = air_model_data)\n\nnew_wind_days <- tibble(Wind = c(5, 10, 15))\n\npredict(ozone_wind_model, newdata = new_wind_days)"
)
```

```{r ex-wind-predictions-solution-ui, echo=FALSE}
uiOutput("ex_wind_predictions_solution_code")
```

## Residuals

A residual is:

$$
\text{observed value} - \text{predicted value}
$$

```{r add-residuals, exercise=TRUE}
model_data_with_residuals <- air_model_data %>%
  mutate(
    predicted_ozone = predict(ozone_temp_model),
    residual = Ozone - predicted_ozone
  )

head(model_data_with_residuals)
```

## Residual Plot

```{r residual-plot, exercise=TRUE}
ggplot(model_data_with_residuals, aes(x = predicted_ozone, y = residual)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Predicted ozone", y = "Residual")
```

## Interpretation Check

In one or two sentences, interpret the slope of the `Ozone ~ Temp`
model in context. Be clear that this model describes an association and
does not prove causation.

```{r ex-slope-interpretation, exercise=TRUE}

```

## Your Turn 3

Fit a model predicting `Ozone` from `Solar.R`. Make a scatterplot with a
fitted line.

```{r ex-solar-model-plot, exercise=TRUE}

```

```{r ex-solar-model-plot-hint-1}
# Fit lm(Ozone ~ Solar.R, data = air_model_data).
# Then map Solar.R to x and Ozone to y in ggplot().
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_solar_model_plot_solution_code",
  "ozone_solar_model <- lm(Ozone ~ Solar.R, data = air_model_data)\n\nsummary(ozone_solar_model)\n\nggplot(air_model_data, aes(x = Solar.R, y = Ozone)) +\n  geom_point() +\n  geom_smooth(method = \"lm\", se = FALSE) +\n  labs(x = \"Solar radiation\", y = \"Ozone\")"
)
```

```{r ex-solar-model-plot-solution-ui, echo=FALSE}
uiOutput("ex_solar_model_plot_solution_code")
```

## Final Task

Choose either `Temp`, `Wind`, or `Solar.R` as the explanatory variable.
Fit a model predicting `Ozone`, then write a short interpretation of the
slope in context.

```{r ex-final-model, exercise=TRUE}

```

```{r ex-final-model-hint-1}
# Example structure:
# my_model <- lm(Ozone ~ Temp, data = air_model_data)
# summary(my_model)
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_final_model_solution_code",
  "my_model <- lm(Ozone ~ Temp, data = air_model_data)\n\nsummary(my_model)\n\n# Interpretation example:\n# For each additional degree Fahrenheit, the model predicts an average\n# increase of about 2.44 ozone units. This is an association in these\n# data, not proof that temperature alone causes ozone to increase."
)
```

```{r ex-final-model-solution-ui, echo=FALSE}
uiOutput("ex_final_model_solution_code")
```

## Submit

```{r, echo=FALSE, context="server"}
encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(ui_before = hash_encoder_ui)
```
