---
title: "Probability: 4 - Functions, Loops, and Probability Calculations"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(shiny)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
tutorial_options(exercise.eval = FALSE)

my_reveal_time <- "2026-06-20 15:00:00"

outcomes <- c(0, 1, 2, 3)
probabilities <- c(0.10, 0.20, 0.40, 0.30)

is_server_context <- function(.envir) {
  inherits(.envir$input, "reactivevalues") &
    inherits(.envir$output, "shinyoutput") &
    inherits(.envir$session, "ShinySession")
}

check_server_context <- function(.envir) {
  if (!is_server_context(.envir)) {
    calling_func <- deparse(sys.calls()[[sys.nframe() - 1]])
    err <- paste0("Function `", calling_func, "` must be called from an Rmd chunk where `context = \"server\"`")
    stop(err, call. = FALSE)
  }
}

encoder_logic <- function(strip_output = FALSE) {
  p <- parent.frame()
  check_server_context(p)
  assign("strip_output", strip_output, envir = p)
  local(
    {
      encoded_txt <- shiny::eventReactive(
        input$hash_generate,
        {
          state <- learnr:::get_tutorial_state()
          shiny::validate(shiny::need(length(state) > 0, "No progress yet."))
          shiny::validate(shiny::need(nchar(input$name) > 0, "No name entered."))
          shiny::validate(shiny::need(nchar(input$studentID) > 0, "Please enter your student ID"))
          user_state <- purrr::map_dfr(state, identity, .id = "label")
          user_state <- dplyr::group_by(user_state, label, type, correct)
          user_state <- dplyr::summarize(
            user_state,
            answer = list(answer),
            timestamp = dplyr::first(timestamp),
            .groups = "drop"
          )
          user_state <- dplyr::relocate(user_state, correct, .before = timestamp)
          user_info <- tibble(
            label = c("student_name", "student_id"),
            type = "identifier",
            answer = as.list(c(input$name, input$studentID)),
            timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "UTC")
          )
          learnrhash::encode_obj(bind_rows(user_info, user_state))
        }
      )
      output$hash_output <- shiny::renderText(encoded_txt())
    },
    envir = p
  )
}

hash_encoder_ui <- {
  shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please enter your identifying information, then click the button below to generate your hash",
    textInput("name", "What's your name?"),
    textInput("studentID", "What is your student ID?"),
    renderText({
      input$caption
    })
  )
}

solution_code_card <- function(code, title = "Solution") {
  tags$div(
    style = paste(
      "margin: 1rem 0 1.25rem;",
      "border: 1px solid #d7e3f3;",
      "border-left: 5px solid #2f6f9f;",
      "border-radius: 8px;",
      "background: #f8fbff;",
      "box-shadow: 0 2px 8px rgba(31, 45, 61, 0.08);",
      "overflow: hidden;"
    ),
    tags$div(
      style = paste(
        "display: flex;",
        "align-items: center;",
        "justify-content: space-between;",
        "gap: 0.75rem;",
        "padding: 0.7rem 0.9rem;",
        "background: #eef6ff;",
        "border-bottom: 1px solid #d7e3f3;"
      ),
      tags$strong(title, style = "color: #1f4e79; font-size: 1rem;"),
      tags$span(
        "released",
        style = paste(
          "padding: 0.15rem 0.5rem;",
          "border-radius: 999px;",
          "background: #ffffff;",
          "color: #3c6478;",
          "font-size: 0.78rem;",
          "font-weight: 600;"
        )
      )
    ),
    tags$pre(
      style = paste(
        "margin: 0;",
        "padding: 0.9rem 1rem;",
        "background: #ffffff;",
        "border: 0;",
        "white-space: pre-wrap;",
        "text-align: left;"
      ),
      tags$code(
        htmltools::HTML(htmltools::htmlEscape(code)),
        style = "display: block; color: #243447; font-size: 0.95rem; line-height: 1.45;"
      )
    )
  )
}

timed_solution <- function(output_id, code, title = "Solution",
                           reveal_time = my_reveal_time,
                           tz = "America/New_York",
                           check_every_ms = 1000) {
  p <- parent.frame()
  check_server_context(p)

  p$output[[output_id]] <- shiny::renderUI({
    shiny::invalidateLater(check_every_ms, p$session)
    release_at <- as.POSIXct(reveal_time, tz = tz)
    if (Sys.time() < release_at) {
      return(NULL)
    }
    solution_code_card(code, title)
  })
}
```

## Welcome

Hello, and welcome to **Functions, Loops, and Probability
Calculations**!

In this tutorial we will use R to turn repeated probability calculations
into reusable code. We will write small functions, simulate repeated
random experiments, and compare simulated results with theoretical
probabilities.

Our learning goals are to:

- write a function with inputs and a returned value;
- check whether a probability distribution is valid;
- compute expected value from outcomes and probabilities;
- use `sample()` and `replicate()` to simulate repeated random outcomes;
- explain why simulated probabilities get more stable as the number of
  repetitions increases.

## Packages

```{r load-packages, exercise=TRUE}
library(dplyr)
library(ggplot2)
```

## Functions for Repeated Calculations

A function is useful when we need to do the same kind of calculation
more than once. For example, probability from counts is:

$$
\text{probability} = \frac{\text{number of favorable outcomes}}{\text{number of possible outcomes}}
$$

```{r probability-function, exercise=TRUE}
prob_from_counts <- function(favorable, total) {
  favorable / total
}

prob_from_counts(favorable = 2, total = 6)
```

## Your Turn 1

Write a function named `percent_from_counts()` that converts favorable
outcomes and total outcomes into a percentage.

```{r ex-percent-function, exercise=TRUE}

```

```{r ex-percent-function-hint-1}
# Start with:
# percent_from_counts <- function(favorable, total) {
#   ...
# }
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_percent_function_solution_code",
  "percent_from_counts <- function(favorable, total) {\n  favorable / total * 100\n}\n\npercent_from_counts(2, 6)"
)
```

```{r ex-percent-function-solution-ui, echo=FALSE}
uiOutput("ex_percent_function_solution_code")
```

## Checking a Probability Distribution

A valid probability distribution has two features:

- every probability is between 0 and 1;
- all probabilities add up to 1.

```{r check-probabilities, exercise=TRUE}
outcomes <- c(0, 1, 2, 3)
probabilities <- c(0.10, 0.20, 0.40, 0.30)

sum(probabilities)
all(probabilities >= 0 & probabilities <= 1)
```

## Your Turn 2

The vector below is supposed to contain probabilities, but it has a
problem. Run the checks, then edit the probabilities so the distribution
is valid.

```{r ex-check-distribution, exercise=TRUE}
probabilities_bad <- c(0.15, 0.25, 0.35, 0.35)

sum(probabilities_bad)
all(probabilities_bad >= 0 & probabilities_bad <= 1)
```

```{r ex-check-distribution-hint-1}
# The current probabilities add to more than 1.
# Change one value so the sum is exactly 1.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_check_distribution_solution_code",
  "probabilities_fixed <- c(0.15, 0.25, 0.35, 0.25)\n\nsum(probabilities_fixed)\nall(probabilities_fixed >= 0 & probabilities_fixed <= 1)"
)
```

```{r ex-check-distribution-solution-ui, echo=FALSE}
uiOutput("ex_check_distribution_solution_code")
```

## Expected Value

Expected value is a weighted average. Outcomes with larger probabilities
count more heavily.

```{r expected-value-example, exercise=TRUE}
outcomes <- c(0, 1, 2, 3)
probabilities <- c(0.10, 0.20, 0.40, 0.30)

sum(outcomes * probabilities)
```

## Your Turn 3

Create a function called `expected_value()` that takes two inputs:
`x` for outcomes and `p` for probabilities. The function should return
the expected value.

```{r ex-expected-value-function, exercise=TRUE}

```

```{r ex-expected-value-function-hint-1}
# Multiply x and p, then add the results.
# expected_value <- function(x, p) {
#   ...
# }
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_expected_value_function_solution_code",
  "expected_value <- function(x, p) {\n  sum(x * p)\n}\n\nexpected_value(outcomes, probabilities)"
)
```

```{r ex-expected-value-function-solution-ui, echo=FALSE}
uiOutput("ex_expected_value_function_solution_code")
```

## Simulating One Random Outcome

We can use `sample()` to draw one outcome according to a probability
distribution.

```{r sample-one-outcome, exercise=TRUE}
set.seed(213)

sample(
  x = outcomes,
  size = 1,
  prob = probabilities
)
```

## Repeating a Simulation

The function `replicate()` repeats the same expression many times.

```{r replicate-simulation, exercise=TRUE}
set.seed(213)

simulated_outcomes <- replicate(
  n = 1000,
  expr = sample(outcomes, size = 1, prob = probabilities)
)

table(simulated_outcomes)
```

## Simulated Probabilities

```{r simulated-probabilities, exercise=TRUE}
simulated_table <- tibble(outcome = simulated_outcomes) %>%
  count(outcome) %>%
  mutate(simulated_probability = n / sum(n))

simulated_table
```

## Compare Simulation and Theory

```{r compare-simulation-theory, exercise=TRUE}
theory_table <- tibble(
  outcome = outcomes,
  theoretical_probability = probabilities
)

left_join(theory_table, simulated_table, by = "outcome")
```

## Your Turn 4

Simulate 100 rolls of a fair six-sided die. Then estimate the
probability of rolling a 6.

```{r ex-die-simulation, exercise=TRUE}

```

```{r ex-die-simulation-hint-1}
# Use sample(1:6, size = 1) inside replicate().
# Then use mean(rolls == 6).
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_die_simulation_solution_code",
  "set.seed(213)\n\nrolls <- replicate(100, sample(1:6, size = 1))\nmean(rolls == 6)"
)
```

```{r ex-die-simulation-solution-ui, echo=FALSE}
uiOutput("ex_die_simulation_solution_code")
```

## Interpretation Check

In one or two sentences, explain why the simulated probability from
100 die rolls may not be exactly equal to the theoretical probability
of rolling a 6.

```{r ex-simulation-interpretation, exercise=TRUE}

```

## Final Task

Improve `expected_value()` so it checks whether the probabilities add to
1. If they do not, the function should stop with an error message.

```{r ex-final-expected-value, exercise=TRUE}

```

```{r ex-final-expected-value-hint-1}
# Use if (...) stop("...")
# round(sum(p), 10) is a useful way to avoid tiny decimal issues.
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_final_expected_value_solution_code",
  "expected_value <- function(x, p) {\n  if (round(sum(p), 10) != 1) {\n    stop(\"Probabilities must sum to 1.\")\n  }\n  sum(x * p)\n}\n\nexpected_value(outcomes, probabilities)"
)
```

```{r ex-final-expected-value-solution-ui, echo=FALSE}
uiOutput("ex_final_expected_value_solution_code")
```

## Submit

```{r, echo=FALSE, context="server"}
encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(ui_before = hash_encoder_ui)
```
