---
title: "Tutorial 2: Data Transformation and Summaries"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(openintro)
library(emo)
library(shiny)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)


my_reveal_time <- "2026-06-20 15:00:00"  # solution reveal time


# Setup for in some exercise chunks ... because each chunk in learnr (shiny app) works independently (not sharing env) 

numbers<-seq(10,60,10)

df <- datasets::airquality
df <- df |> drop_na()




# Hash generation helpers
# Should ideally be loaded from the imstutorials package when it exists
is_server_context <- function(.envir) {
  # We are in the server context if there are the follow:
  # * input - input reactive values
  # * output - shiny output
  # * session - shiny session
  #
  # Check context by examining the class of each of these.
  # If any is missing then it will be a NULL which will fail.
  
  inherits(.envir$input, "reactivevalues") &
    inherits(.envir$output, "shinyoutput") &
    inherits(.envir$session, "ShinySession")
}

check_server_context <- function(.envir) {
  if (!is_server_context(.envir)) {
    calling_func <- deparse(sys.calls()[[sys.nframe() - 1]])
    err <- paste0("Function `", calling_func, "`", " must be called from an Rmd chunk where `context = \"server\"`")
    stop(err, call. = FALSE)
  }
}
encoder_logic <- function(strip_output = FALSE) {
  p <- parent.frame()
  check_server_context(p)
  # Make this var available within the local context below
  assign("strip_output", strip_output, envir = p)
  # Evaluate in parent frame to get input, output, and session
  local(
    {
      encoded_txt <- shiny::eventReactive(
        input$hash_generate,
        {
          # shiny::getDefaultReactiveDomain()$userData$tutorial_state
          state <- learnr:::get_tutorial_state()
          shiny::validate(shiny::need(length(state) > 0, "No progress yet."))
          shiny::validate(shiny::need(nchar(input$name) > 0, "No name entered."))
          shiny::validate(shiny::need(nchar(input$studentID) > 0, "Please enter your student ID"))
          user_state <- purrr::map_dfr(state, identity, .id = "label")
          user_state <- dplyr::group_by(user_state, label, type, correct)
          user_state <- dplyr::summarize(
            user_state,
            answer = list(answer),
            timestamp = dplyr::first(timestamp),
            .groups = "drop"
          )
          user_state <- dplyr::relocate(user_state, correct, .before = timestamp)
          user_info <- tibble(
            label = c("student_name", "student_id"),
            type = "identifier",
            answer = as.list(c(input$name, input$studentID)),
            timestamp = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "UTC")
          )
          learnrhash::encode_obj(bind_rows(user_info, user_state))
        }
      )
      output$hash_output <- shiny::renderText(encoded_txt())
    },
    envir = p
  )
}

hash_encoder_ui <- {
  shiny::div("If you have completed this tutorial and are happy with all of your", "solutions, please enter your identifying information, then click the button below to generate your hash", textInput("name", "What's your name?"), textInput("studentID", "What is your student ID?"), renderText({
    input$caption
  }), )
}





solution_code_card <- function(code, title = "Solution") {
  tags$div(
    style = paste(
      "margin: 1rem 0 1.25rem;",
      "border: 1px solid #d7e3f3;",
      "border-left: 5px solid #2f6f9f;",
      "border-radius: 8px;",
      "background: #f8fbff;",
      "box-shadow: 0 2px 8px rgba(31, 45, 61, 0.08);",
      "overflow: hidden;"
    ),
    tags$div(
      style = paste(
        "display: flex;",
        "align-items: center;",
        "justify-content: space-between;",
        "gap: 0.75rem;",
        "padding: 0.7rem 0.9rem;",
        "background: #eef6ff;",
        "border-bottom: 1px solid #d7e3f3;"
      ),
      tags$strong(
        title,
        style = "color: #1f4e79; font-size: 1rem;"
      ),
      tags$span(
        "released",
        style = paste(
          "padding: 0.15rem 0.5rem;",
          "border-radius: 999px;",
          "background: #ffffff;",
          "color: #3c6478;",
          "font-size: 0.78rem;",
          "font-weight: 600;"
        )
      )
    ),
    tags$pre(
      style = paste(
        "margin: 0;",
        "padding: 0.9rem 1rem;",
        "background: #ffffff;",
        "border: 0;",
        "border-radius: 0;",
        "white-space: pre-wrap;",
        "text-align: left;"
      ),
      tags$code(
        htmltools::HTML(htmltools::htmlEscape(code)),
        style = paste(
          "display: block;",
          "margin: 0;",
          "padding: 0;",
          "color: #243447;",
          "font-size: 0.95rem;",
          "line-height: 1.45;",
          "text-align: left;",
          "text-indent: 0;"
        )
      )
    )
  )
}



timed_solution <- function(output_id, code, title = "Solution",
                           reveal_time = my_reveal_time,
                           tz = "America/New_York",
                           check_every_ms = 1000) {
  p <- parent.frame()
  check_server_context(p)

  p$output[[output_id]] <- shiny::renderUI({
    shiny::invalidateLater(check_every_ms, p$session)

    release_at <- as.POSIXct(reveal_time, tz = tz)
    if (Sys.time() < release_at) {
      return(NULL)
    }

    solution_code_card(code, title)
  })
}
```

## Welcome

Hello, and welcome to **Data Transformation and Summaries**!

In this tutorial, we will use R to load data, inspect variables, create
new variables, filter rows, remove missing values, and make contingency
tables. Together with Tutorial 1, this tutorial prepares you for
**Lab 2**.

Our learning goals are to:

- use `dplyr` verbs such as `select()`, `filter()`, `mutate()`, and
  `summarise()`;
- create categorized variables from numerical variables;
- remove missing values with `filter()` and `drop_na()`;
- make contingency tables with `group_by()`, `summarise()`, and
  `pivot_wider()`.

## Packages

This lesson uses three packages:

- **dplyr** for data transformation
- **readr** for reading data files
- **tidyr** for reshaping data and removing missing values

```{r load-packages, exercise=TRUE}
library(dplyr)
library(readr)
library(tidyr)
```

## New York Air Quality Data

The New York air quality data contain daily air quality measurements
from New York City from May to September 1973. The full data set has 153
observations and 6 variables.

Each row is one day. The variables include ozone level, solar radiation,
wind speed, temperature, month, and day.

You can read more about the original data description here:
<https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/airquality.html>

## Import the Data

For the interactive exercises, we have already loaded the complete cases
from the air quality data into an object named `df`. Print the data.

```{r import-airquality-setup, include=FALSE}
df <- datasets::airquality
df <- df %>% drop_na()
```

```{r import-airquality, exercise=TRUE, exercise.setup="import-airquality-setup"}
print(df)
```

## Inspect the Data

Before analyzing a data set, it is helpful to look at its structure,
variable names, and first few rows.

```{r inspect-airquality, exercise=TRUE, exercise.setup="import-airquality-setup"}
glimpse(df)
names(df)
head(df)
```

## What Did You See in the Output?

```{r question1, echo=FALSE}
quiz(
  question(
    "How many rows are in `df`?",
    answer("6"),
    answer("Ozone"),
    answer("10"),
    answer("111", correct = TRUE),
    allow_retry = TRUE
  )
)
```

## The Pipe Operator

The pipe operator sends the result of one step into the next step. You
can read it as "then."

`df %>% filter(Month == 5)` means "start with `df`, then select all rows
where `Month` equals 5."

```{r pipe-filter-example, exercise=TRUE, exercise.setup="import-airquality-setup"}
df %>% 
  filter(Month == 5)
```

The native R pipe `|>` works similarly.

```{r native-pipe-example2, exercise=TRUE, exercise.setup="import-airquality-setup"}
df |> 
  select(Ozone, Wind, Month) |>
  filter(Month == 5)
```

Pipe operators can be combined with assignment to create a new object.
For example, if you want to select the `Ozone`, `Wind`, and `Month`
columns, keep only observations from May, and assign the result to
`df2`, you can write:

```{r native-pipe-example3, exercise=TRUE, exercise.setup="import-airquality-setup"}
df2 <- df |> 
  select(Ozone, Wind, Month) |>
  filter(Month == 5)

print(df2)
```

## `dplyr` Data Manipulation Functions

Here is a summary of commonly used `dplyr` functions. It is helpful to
think about whether each function works mainly with rows, columns, or
groups of rows.

- Rows:

  - [`filter()`](https://dplyr.tidyverse.org/reference/filter.html)
    chooses rows based on column values.

    Usage: `df |> filter(Month == 5)`: keeps only the rows from May.

  - [`slice()`](https://dplyr.tidyverse.org/reference/slice.html)
    chooses rows based on location.

    Usage: `df |> slice(1:5)`: keeps only rows 1 through 5.

  - [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html)
    changes the order of the rows.

    Usage: `df |> arrange(Temp)`: sorts the rows from lowest
    temperature to highest temperature.

- Columns:

  - [`select()`](https://dplyr.tidyverse.org/reference/select.html)
    chooses which columns to keep.

    Usage: `df |> select(Ozone, Temp, Month)`: keeps only the `Ozone`,
    `Temp`, and `Month` columns.

  - [`rename()`](https://dplyr.tidyverse.org/reference/rename.html)
    changes the name of columns.

    Usage: `df |> rename(Temperature = Temp)`: changes the column name
    `Temp` to `Temperature`.

  - [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html)
    changes the values of columns and creates new columns.

    Usage: `df |> mutate(Temp_C = (Temp - 32) * 5 / 9)`: creates a new
    Celsius temperature column named `Temp_C`.

  - [`relocate()`](https://dplyr.tidyverse.org/reference/relocate.html)
    changes the order of the columns.

    Usage: `df |> relocate(Month, Day)`: moves the `Month` and `Day`
    columns to the front.

- Groups of rows:

  - [`summarise()`](https://dplyr.tidyverse.org/reference/summarise.html)
    collapses a group into a single row.

    Usage: `df |> summarise(avg_temp = mean(Temp))`: calculates one
    summary value, the average temperature.

## Creating a New Column

Use `mutate()` to create a new column or modify an existing column.

For example, we can create a new temperature category named
`Temp_Category` where the temperature is categorized as High if the
temperature is greater than 80 and Moderate otherwise.

```{r mutate-temp-category, exercise=TRUE, exercise.setup="import-airquality-setup"}
df2 <- df %>%
  mutate(Temp_Category = ifelse(Temp > 80, "High", "Moderate"))

head(df2)
```

## Contingency Table 1

A contingency table shows counts for combinations of categorical
variables.

First, we create a categorized ozone variable. It is `High` if the ozone
level is greater than 42 and `Low` otherwise.

```{r contingency-table-ozone, exercise=TRUE, exercise.setup="import-airquality-setup"}
  df %>%
  mutate(Ozone_Level = ifelse(Ozone > 42,
                              "High", "Low"))

```

Then we count the number of days in each combination of `Month` and
`Ozone_Level`.

```{r contingency-table-ozone2, exercise=TRUE, exercise.setup="import-airquality-setup"}

  df %>%
  mutate(Ozone_Level = ifelse(Ozone > 42,
                              "High", "Low"))%>%
  group_by(Month, Ozone_Level) 

```

After that, we create a new count variable.

```{r contingency-table-ozone3, exercise=TRUE, exercise.setup="import-airquality-setup"}

  df %>%
  mutate(Ozone_Level = ifelse(Ozone > 42,
                              "High", "Low"))%>%
  group_by(Month, Ozone_Level) %>%
  summarise(Count = n())

```

Now we can use the `Count` column as the entries in the table by calling
`pivot_wider()`.

```{r contingency-table-ozone4, exercise=TRUE, exercise.setup="import-airquality-setup"}

contingency_table <- df %>%
  mutate(Ozone_Level = ifelse(Ozone > 42,
                              "High", "Low"))%>%
  group_by(Month, Ozone_Level) %>%
  summarise(Count = n()) %>%
  pivot_wider(names_from = Ozone_Level, values_from = Count)

contingency_table

```

## Contingency Table 2

Next, let's create a categorized version of `Solar.R` with three
categories.

```         
Solar.R <= 115       : Low
115 < Solar.R <= 185 : Medium
185 < Solar.R        : High
```

```{r contingency-table-solar-setup, include=FALSE}
df4 <- df %>%
  mutate(Ozone_Level = ifelse(Ozone > median(Ozone),
                              "High", "Low"))
```

First, create a categorical ozone variable named `Ozone_Level`.

```{r contingency-table-solar, exercise=TRUE, exercise.setup="contingency-table-solar-setup"}

df4 <- df %>%
  mutate(Ozone_Level = ifelse(Ozone > median(Ozone),
                              "High", "Low"))
head(df4)
```

Using `df4`, we can create a new categorical variable based on
`Solar.R`.

```{r contingency-table-solar2, exercise=TRUE, exercise.setup="contingency-table-solar-setup"}
df4 %>%
  mutate(solar_cat = case_when(
    Solar.R <= 115 ~ "Low",
    Solar.R > 115 & Solar.R <= 185 ~ "Medium",
    Solar.R > 185 ~ "High"
  )) 
```

Then, using `solar_cat` and `Ozone_Level`, we can create a count
variable with the `n()` function.

```{r contingency-table-solar3, exercise=TRUE, exercise.setup="contingency-table-solar-setup"}
df4 %>%
  mutate(solar_cat = case_when(
    Solar.R <= 115 ~ "Low",
    Solar.R > 115 & Solar.R <= 185 ~ "Medium",
    Solar.R > 258 ~ "High"
  )) %>%
  group_by(solar_cat, Ozone_Level) %>%
  summarise(Count = n()) 

```

Finally, use `pivot_wider()` to create a contingency table.

```{r contingency-table-solar4, exercise=TRUE, exercise.setup="contingency-table-solar-setup"}
contingency_table2 <- df4 %>%
  mutate(solar_cat = case_when(
    Solar.R <= 115 ~ "Low",
    Solar.R > 115 & Solar.R <= 185 ~ "Medium",
    Solar.R > 258 ~ "High"
  )) %>%
  group_by(solar_cat, Ozone_Level) %>%
  summarise(Count = n()) %>%
  pivot_wider(
    names_from = Ozone_Level,
    values_from = Count
  )

print(contingency_table2)
```


## Your Turn 1

### Create Celsius Temperature

In the air quality data, create a new column named `Temp_C` from `Temp`.
Use this formula:

$$C = (F - 32) \times \frac{5}{9}$$

```{r ex-temp-c, exercise=TRUE}
df <- datasets::airquality
df <- df |> drop_na()


```

```{r ex-temp-c-hint-1}
# Use mutate() to create Temp_C.
```

```{r ex-temp-c-hint-2}
# The right side should be the Celsius formula:
# (Temp - 32) * 5 / 9
```

```{r, context="server", echo=FALSE}
timed_solution(
  "ex_temp_c_solution_code",
  r"(
df <- datasets::airquality
df <- df |> drop_na()

df_new1 <- df %>%
  mutate(Temp_C = (Temp - 32) * 5 / 9)

print(df_new1)
)"
)
```

```{r ex-temp-c-solution-ui, echo=FALSE}
uiOutput("ex_temp_c_solution_code")
```


## Your Turn 2

### Categorize Wind and Ozone

The Beaufort wind scale categorizes wind speed in miles per hour:

```         
Calm: less than 1 mph
Light Air: 1-4 mph
Light Breeze: 4-7 mph
Gentle Breeze: 7-12 mph
Moderate Breeze: 12-18 mph
Fresh Breeze: 18-24 mph
Strong Breeze: greater than 24 mph
```

Also categorize ozone level into two levels:

```         
Low: less than or equal to the median ozone level
High: greater than the median ozone level
```

### Step 1: Create an Ozone Category

```{r ex-ozone-category, exercise=TRUE}
df 
  
```

```{r ex-ozone-category-hint-1}
# Start with creating new categorical variable `Ozone_cat` 
```

```{r ex-ozone-category-hint-2}
# You can go back to `Contingency Table 1 section and come back here
```

```{r ex-ozone-category-hint-3}
# To obtain the median, you can use median(Ozone)
```

### Step 2: Create a Wind Category

```{r ex-wind-category, exercise=TRUE}
```

```{r ex-wind-category-hint-1}
# Start with creating new categorical variable `Wind_cat` 
```

```{r ex-wind-category-hint-2}
# You can go back to `Contingency Table 2 section and come back here
```

```{r ex-wind-category-hint-3}
# For more than 2 categories, you can use case_when() within mutate() 
```

### Step 3: Make a Contingency Table

```{r ex-wind-ozone-table, exercise=TRUE}
```

```{r ex-wind-ozone-table-hint-1}
# Use group_by() and pivot_wider()
```


```{r, context="server", echo=FALSE}
timed_solution(
  "ex_wind_ozone_solution_code",
  r"(
  #Step 1
df_categorized <- df %>%
  mutate(Ozone_cat = case_when(
    Ozone <= median(Ozone) ~ "Low",
    Ozone > median(Ozone) ~ "High"
  ))

 #Step 2
df_categorized2 <- df_categorized %>%
  mutate(Wind_cat = case_when(
    Wind <= 1 ~ "Calm",
    Wind > 1 & Wind <= 4 ~ "Light Air",
    Wind > 4 & Wind <= 7 ~ "Light Breeze",
    Wind > 7 & Wind <= 12 ~ "Gentle Breeze",
    Wind > 12 & Wind <= 18 ~ "Moderate Breeze",
    Wind > 18 & Wind <= 24 ~ "Fresh Breeze",
    Wind > 24 ~ "Strong Breeze"
  ))

 #Step 3
my_contingency_tbl <- df_categorized2 %>%
  group_by(Ozone_cat, Wind_cat) %>%
  summarise(Count = n()) %>%
  pivot_wider(names_from = Wind_cat, values_from = Count)

my_contingency_tbl
)"
)
```

```{r ex-wind-ozone-solution-ui, echo=FALSE}
uiOutput("ex_wind_ozone_solution_code")
```

## Submit

```{r, echo=FALSE, context="server"}
encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(ui_before = hash_encoder_ui)
```
